Type theory in type theory with single substitutions

### 2024-09-09 types_2024_postproc/p.tex
stuff

\newcommand{\ra}{\rightarrow}
\newcommand{\Set}{\mathsf{Set}}
\newcommand{\Ty}{\mathsf{Ty}}
\newcommand{\Tm}{\mathsf{Tm}}
\newcommand{\p}{\mathsf{p}}
\newcommand{\q}{\mathsf{q}}
\newcommand{\ext}{\mathop{\triangleright}}
\newcommand{\N}{\mathbb{N}}
\newcommand{\lam}{\mathsf{lam}}
\newcommand{\app}{\mathsf{app}}
\newcommand{\U}{\mathsf{U}}
\newcommand{\El}{\mathsf{El}}
\newcommand{\cd}{\mathsf{c}}

@@

In this paper, we define the SSC and show how it can be extended with
$\Pi$ types and a hierarchy of Coquand-universes
\cite{coquandUniverse}. The rules for type formers are the same as for
CwFs, the difference is only in the substitution calculus part. More
generally, every second-order generalised algebraic theory (SOGAT) has
a parallel and a single substition-based first order presentation
\cite{DBLP:conf/fscd/KaposiX24}. Here we prove that the syntaxes of
the parallel and the single substitution calculi are isomorphic for
the following SOGAT (we use implicit arguments and $\cong$ denotes
isomorphism):
\begin{alignat*}{10}
  & \Ty && : \N\ra\Set \\
  & \Tm && : \Ty\,i\ra\Set \\
  & \Pi && : (A:\Ty\,i)\ra(\Tm\,A\ra\Ty\,i)\ra\Ty\,i \\
  & \app && : \Tm\,(\Pi\,A\,B)\cong((a:\Tm\,A)\ra\,\Tm\,(B\,a)): \lam \\
  & \U && : (i:\N)\ra\Ty\,(1+i) \\
  & \El && : \Tm\,(\U\,i)\cong\Ty\,i : \cd
\end{alignat*}
We conjecture that the methods of in this paper can be used to show
that for any SOGAT, the two first-order presentations give isomorphic
syntaxes.

TODO: say something about alpha-normalisation.

To summarise, our contributions are the following:
\begin{itemize}
\item A new generalised algebraic presentation of type theory in the
  form of a minimalistic single substitution calculus.
\item The $\alpha$-normalisation technique which is useful for working
  with explicit substitution calculi.
\item The proof of equivalence with the parallel, CwF-based syntax.
\end{itemize}

@@

In Section \ref{sec:ssc}, we introduce

We first show how to get rid of explicit substitutions from the syntax
using the technique of $\alpha$-normalisation (Section
\ref{sec:alpha}). Then we show how to derive the rules of a CwF from
our syntax (Section \ref{sec:cwf}). Finally, we show that our syntax
is isomorphic to the CwF-style syntax (Section \ref{sec:iso}). We list
related work and conclude in Section \ref{sec:conclusion}.

We only require a working knowledge of an implementation of type
theory such as Agda. Our metatheory in which this paper is written is
thus type theory, and everything is formalised in Agda.

@@

Algebraic formalisations of type theory: Brunerie
\cite{initiality-agda} \cite{DBLP:conf/popl/AltenkirchK16},
normalisation, shallow embedding canonicity, gluing. Low level
formalisations: Abel, Loic, metacoq. They can be seen as unfolding the
more abstract thing in a (weak) setoid model.

@@

@@

\section{Deriving parallel substituions}

@@

\section{Isomorphism}

>>>>

\newcommand{\ra}{\rightarrow}
\newcommand{\Set}{\mathsf{Set}}
\newcommand{\Ty}{\mathsf{Ty}}
\newcommand{\Tm}{\mathsf{Tm}}
\newcommand{\Con}{\mathsf{Con}}
\newcommand{\Sub}{\mathsf{Sub}}
\newcommand{\p}{\mathsf{p}}
\newcommand{\q}{\mathsf{q}}
\newcommand{\ext}{\mathop{\triangleright}}
\newcommand{\N}{\mathbb{N}}
\newcommand{\lam}{\mathsf{lam}}
\newcommand{\app}{\mathsf{app}}
\newcommand{\U}{\mathsf{U}}
\newcommand{\El}{\mathsf{El}}
\newcommand{\cd}{\mathsf{c}}
\newcommand{\blank}{\mathord{\hspace{1pt}\text{--}\hspace{1pt}}} %from the book

@@

We define the parallel substitution calculus operations by induction
on the SSC-syntax. One of our constructors is the instantiation
operation of single substitutions: $\blank[\blank] :
\Ty\,\Gamma\ra\Sub\,\Delta\,\Gamma\ra\Ty\,\Delta$, and similarly for
terms. When doing induction on terms, we always have to provide a
method for these constructors, but this can be difficult as
substitutions don't compose. This is why we derive a new induction
principle which works on $\alpha$-normal forms. Just as $\beta$-normal
forms don't distinguish between $\beta$-equal terms, $\alpha$-normal
forms don't include explicit substitutions, all the variables are
instantiated. $\alpha$-normal forms are not usual normal forms in the
sense that they are still quotiented by the computation/uniqueness
rules like $\beta$/$\eta$ for function space.

In this paper, we define the SSC and extend it with
$\Pi$ types and a hierarchy of Coquand-universes
\cite{coquandUniverse}. The rules for type formers are the same as for
CwFs, the difference is only in the substitution calculus part. More
generally, every second-order generalised algebraic theory (SOGAT) has
a parallel and a single substition-based first order presentation
\cite{DBLP:conf/fscd/KaposiX24}. Here we prove that the syntaxes of
the parallel and the single substitution calculi are isomorphic for
the following SOGAT (we use implicit arguments and $\cong$ denotes
isomorphism):
\begin{alignat*}{10}
  & \Ty && : \N\ra\Set \\
  & \Tm && : \Ty\,i\ra\Set \\
  & \Pi && : (A:\Ty\,i)\ra(\Tm\,A\ra\Ty\,i)\ra\Ty\,i \\
  & \app && : \Tm\,(\Pi\,A\,B)\cong((a:\Tm\,A)\ra\,\Tm\,(B\,a)): \lam \\
  & \U && : (i:\N)\ra\Ty\,(1+i) \\
  & \El && : \Tm\,(\U\,i)\cong\Ty\,i : \cd
\end{alignat*}
We conjecture that the methods of in this paper can be used to show
that for any SOGAT, the two first-order presentations give isomorphic
syntaxes.

To summarise, our contributions are the following:
\begin{itemize}
\item A new generalised algebraic presentation of type theory in the
  form of a minimalistic single substitution calculus (SSC).
\item The $\alpha$-normalisation technique which is useful for working
  with explicit substitution calculi.
\item The proof of equivalence of the SSC-syntax with the CwF-based
  syntax.
\end{itemize}

@@

In Section \ref{sec:ssc}, we introduce the single substitution
calculus only relying on a working knowledge of an implementation of
type theory such as Agda, no prior experience with the metatheory of
type theory is required. We first show how to get rid of explicit
substitutions from the syntax using the technique of
$\alpha$-normalisation (Section \ref{sec:alpha}), and we use this to
prove several admissible equations. Then we show how to derive the
rules of a CwF from our syntax (Section \ref{sec:cwf}). Finally, we
show that our syntax is isomorphic to the CwF-style syntax (Section
\ref{sec:iso}). We conclude in Section \ref{sec:conclusion}.

@@

Most formalisation of type theory use low-level

Algebraic formalisations of type theory: Brunerie

\cite{initiality-agda} \cite{DBLP:conf/popl/AltenkirchK16},
normalisation, shallow embedding canonicity, gluing. Low level
formalisations: Abel, Loic, metacoq. They can be seen as unfolding the
more abstract thing in a (weak) setoid model.

@@

Our metatheory in which this paper is
written is thus type theory, and everything is formalised in Agda.

@@

\section{$\alpha$-normalisation}
\label{sec:alpha}

$\alpha$-normal forms could also be called substitution normal forms.

\section{Isomorphism}
\label{sec:cwf}

@@

TODO: show that the SSC operations are derivable from CwF.

\section{Isomorphism}
\label{sec:iso}
